\documentclass[DIV=15, fleqn, oneside, 8pt]{scrartcl}

% --------------------------------------------------
% Page appearance (subtle warm tone)
% --------------------------------------------------
\usepackage{xcolor}
\pagecolor{yellow!3}        
\color{black}    

% --------------------------------------------------
% Layout and spacing
% --------------------------------------------------
\KOMAoptions{parskip=half}  % space between paragraphs (no indentation feel)
\setlength{\parindent}{0pt} % remove paragraph indentation
\setlength{\mathindent}{0pt}% left-align equations (fleqn tweak)

% --------------------------------------------------
% Math packages
% --------------------------------------------------
\usepackage{amsmath, amssymb}
\AtBeginDocument{\setlength{\mathindent}{0pt}}

% --------------------------------------------------
% Theorem styling (clean, minimal)
% --------------------------------------------------
\usepackage{amsthm, tikz}

\newtheoremstyle{clean}
  {6pt}        % Space above
  {6pt}        % Space below
  {}% Body font
  {}           % Indent amount
  {\bfseries}  % Head font (bold)
  {.}          % Punctuation after heading
  {0.5em}      % Space after heading
  {}

\theoremstyle{clean}
\newtheorem{question}{Q} % auto-numbered questions

% --------------------------------------------------
% Solution environment
% --------------------------------------------------
\newenvironment{solution}
{\par\textit{Sol.} }     % italic label
{\par\medskip\hrule\bigskip}

% --------------------------------------------------
% Title (no date)
% --------------------------------------------------
\title{Questions on Probability}
\date{}                    % removes date

% --------------------------------------------------
\begin{document}
\maketitle

\newpage
\section*{Symmetry \& Total Probability}
\begin{question}
Two gamblers A and B have $n+1$ and $n$ fair coins respectively. If both toss
all their coins, find the probability that the number of heads of A is greater than the
number of heads of B.
\end{question}
\begin{solution}
Define events: \\
$E$: $A$ has more heads in the first $n$ tosses \\
$F$: $B$ has more heads in the first $n$ tosses \\
$G$: $A$ and $B$ have equal heads in the first $n$ tosses \\
$H$: $A$ gets a head in $(n+1)^{th}$ toss

By law of total probability, $P(E) + P(F) + P(G) = 1$. By symmetry, $P(E) = P(F)$, so:
\begin{equation}
2P(E) + P(G) = 1
\end{equation}

We require the probability that $A$ has more heads than $B$ after $n+1$ tosses. This occurs if $E$ happens (regardless of $H$) OR if $G$ happens AND $H$ happens:
\[ P(\text{Result}) = P(E) + P(G)P(H) \]

Since $P(H) = \frac{1}{2}$, we have:
\[ P(\text{Result}) = P(E) + \frac{1}{2}P(G) = \frac{2P(E) + P(G)}{2} \]

Substituting from (1), we get $P(\text{Result}) = \frac{1}{2}$
\end{solution}

% ---

\begin{question}
A, B and C toss a fair coin and get their first heads in $N_A$, $N_B$ and $N_C$ tosses
respectively.
Find $P(N_A < N_B < N_C)$.
\end{question}
\begin{solution}
Let D be the event that $N_A \neq N_B \neq N_C$. By symmetry, each of the 6 permutations under the
event D are equally likely:
\[P(N_A < N_B < N_C) = \frac{1}{6} \times P(D)\]
\begin{equation}
  P(D) = 1 - 3 \times P(two\:are\:same) - P(three\:are\:same)
\end{equation}
$N_A$, $N_B$ and $N_C$ follow the geometric distribution, $(1/2)^n$ where n=1,2... so,
\[P(two\:are\:same) = \sum_{k = 1}^{n}{(\frac{1}{2})^{2n}(1 - (\frac{1}{2})^n)} = 
\sum_{k = 1}^{n}{(\frac{1}{2})^{2n}} - \sum_{k = 1}^{n}{(\frac{1}{2})^{3n}}
= \frac{1}{3} - \frac{1}{7} = \frac{4}{21}\]\[
P(three\:are\:same) = \sum_{k = 1}^{n}{(\frac{1}{2})^{3n}} = \frac{1}{7}\]
Substituting the above values in (2),
\[
P(D) = 1 - 3 \times \frac{4}{21} - \frac{1}{7} = \frac{2}{7} \text{ hence, }
P(N_A < N_B < N_C) = \frac{1}{6} \times \frac{2}{7} = \frac{1}{21}
\]
\end{solution}

\begin{question}
There are $2^n$ players in a knockount style chess tournament, numbered 1,2,3 . . . $2^n$. 
Player numbered $i$ always wins against player number $j\:\forall j > i$.
What is the probability that the $k^{th}$ best player comes second?
\end{question}
\begin{solution}
W.L.O.G, consider this tournament as a pool style tournament. Let $N = 2^n$ and consider two
pools of $N/2$ players each. The tournament progresses separately in both pools and the final is 
played between the champions from each. Let $P_k$ denote the required probability.

It is evident that $P_k = 0 \: \forall k > N/2$ as this $k^{th}$ best player can never emerge as the
champion from a pool, which is the requisite condition for being second.

Also, $P_1$ = 0 as he would always come first.

When $1 < k \leq N/2$, the experiment can be thought of as follows: consider any permutation of $N$ 
players in which the first $N/2$ belong to pool 1 and the rest to pool 2. $P_k$ is the probability 
of the event where players 1,2,..$k-1$ remain in one pool and the $k^{th}$ player remains is the other:
\[
P_k = 1 \times \frac{(N/2) - 1}{N - 1} \times \frac{(N/2) - 2}{N - 2}\:.\:.\:.\:\times
\frac{(N/2) - (k - 2)}{N - (k - 2)} \times \frac{(N/2)}{N - (k-1)}
\]

e.g. when $N = 16$ and $k = 2$, the answer is $\frac{8}{15}$ and when $k=3$, the answer
is $\frac{7}{15} \times \frac{8}{14} = \frac{4}{15}$
\end{solution}

\begin{question}
A fair 40-sided and 60-sided dice are rolled. What is the probability that the number on the 60-sided
die is larger than the 40-sided die?
\end{question}
\begin{solution}
Let S be the required event. Condition S on two events:

$S_1$: the 60-sided die shows a number between $[1, 40]$

$S_2$: the 60-sided die shows a number between $[41, 60]$
\begin{equation}
P(S) = P(S_1)P(S | S_1) + P(S_2)P(S | S_2)
\end{equation}
$P(S | S_2) = 1$. For $P(S | S_1)$, there are 3 possible events:  60-sided die shows a larger
number OR 40-sided die shows a larger number OR both show the same number. By Symmetry:

P(60-sided $>$ 40-sided $|$ $S_1$) = P(40-sided $>$ 60-sided $|$ $S_1$) = x (say)

P(both same) = $\frac{1}{4}$

By law of total probability,
\[2x + \frac{1}{40} = 1 \implies x = \frac{39}{80}\]
Substituting in (3), we get:
\[P(S) = \frac{40}{60} \times \frac{39}{80} + \frac{20}{60} \times 1\ = \frac{79}{120}\]
\end{solution}
\newpage
\section*{Counting By Reflection}
\begin{question}
38 people waiting to see a tennis match. The entrance fee is \$5 to enter.
19 of the people only have \$10 bills, 
while the other 19 only have \$5 bills. The cashier currently has no change.
If the 38 people arrange themselves in line completely at random, 
what is the probability that all of the people can successfully purchase a ticket without 
reordering in line?
\end{question}
\begin{solution}
Consider the $P \times Q$ grid in which the x-axis denotes a \$5 bill and the y-axis denotes
a \$10 bill.

We want to count all paths from $(0, 0)$ to $(P, Q)$ where the y co-ordinate never exceeds the
x co-ordinate i.e. 
\begin{equation}
  \text{\#(total paths) $-$ \#(bad paths : paths that touch or cross the line $y = x + 1$)}
\end{equation}

Let Q' be a bad path. The portion of Q' between $(0, 0)$ and the first point of contact
with the line $y = x + 1$ can be reflected about the same line
to give a path from $(-1, 1)$ to $(P, Q)$.

% diagram
\begin{center}
\begin{tikzpicture}[scale=0.6]
    % 1. Setup Grid and Axes
    \draw[step=1, gray!20, thin] (-2,-1) grid (7,7);
    \draw[->, thick] (-2,0) -- (8,0) node[right] {+5s ($x$)};
    \draw[->, thick] (0,-1) -- (0,8) node[above] {+10s ($y$)};

    % 2. Draw Boundaries
    % Blue = Allowed line y=x. Red = Forbidden boundary y=x+1.
    \draw[blue, ultra thin] (0,0) -- (7,7) node[right] {$y = x$};
    \draw[red, ultra thick, dashed] (-2,-1) -- (6,7) node[pos=0.9, sloped, above] {Forbidden: $y = x + 1$};

    % 3. The "Bad" Path (from (0,0) to (6,6))
    % This path touches y=x+1 at (0,1).
    % Path breakdown: (0,0) -> (0,1) -> (1,1) -> (1,4) -> (3,4) -> (3,5) -> (5,5) -> (5,6) -> (6,6)
    \draw[orange, ultra thick] (0,0) -- (0,1) -- (1,1) -- (1,4) -- (3,4) -- (3,5) -- (5,5) -- (5,6) -- (6,6);
    
    % 4. First Contact Point (where bad path hits y=x+1)
    \filldraw[red] (0,1) circle (5pt) node[above left] {Contact};

    % 5. The "Reflected" Segment (from (-1,1) to (0,1))
    % Reflection maps (0,0) to (-1,1).
    % Segment from (-1,1) to (0,1) replaces the (0,0) to (0,1) segment.
    \draw[purple, ultra thick, dotted] (-1,1) -- (0,1);

    % 6. Label Critical Points
    \filldraw[black] (0,0) circle (4pt) node[below right] {$A(0,0)$};
    \filldraw[purple] (-1,1) circle (4pt) node[left] {$A'(-1,1)$};
    \filldraw[black] (6,6) circle (4pt) node[right] {$B(6,6)$};

\end{tikzpicture}
\end{center}

Similarly, any path from $(-1, 1)$ to $(P, Q)$ can be converted to a bad path.

There is a one-one correspondence between the set of bad paths and the set of paths from
$(-1, 1)$ to $(P, Q)$.

Number of paths from $(-1, 1)$ to $(P, Q)$ is equal to ${P + Q \choose Q - 1} = \text{\#(bad paths)}$

Substituting in (3),
\[
\text{\#(good paths)} = {P+Q \choose Q} - {P+Q \choose Q-1} = \frac{1}{Q+1} {P+Q \choose Q}
\]

Dividing by the total number of paths, we get the required probability = $\frac{1}{Q+1}$.

Since here $P=Q=19$, the answer is $\frac{1}{20}$.
\end{solution}

\newpage
\section*{Optimal Stopping Problems}
\begin{question}
Consider a dice game where upto three rolls of a fair 6 sided die are allowed. The first roll
is free. A player can choose to accept the reward as the face value or forfeit it, pay \$1 and roll
again. The same rule applies for the second roll. In the third roll, there is no choice but to
accept the reward. What is the optimal strategy? What is the expected reward under this optimal
strategy?
\end{question}
\begin{solution}
\textbf{This is an optimal stopping problem where we need to decide the stopping rule for the 
first two rolls.}


The stopping decision for any roll should depend on the future potential, as there is no reward
from the past rolls. Due to this observation, we need to approach this problem via
\textbf{backward induction}.

Let $E_1 \: E_2 \: and\: E_3$ denote the potential reward from starting the game at rolls 1, 2 and
3 respectively.

$E_3 = 3.5 - 1$ = 2.5 where 3.5 is the expected reward and 1 for the fee of rolling the die a 
third time.

This means, at roll 2, we must not accept anything less than 2.5. Stopping decision for roll 2 is
to stop if we roll 3, 4, 5 or 6, or, roll again otherwise. This gives:

\[
E_2 = (\frac{2}{6} \times 2.5) + (\frac{4}{6} \times \frac{3 + 4 + 5 + 6}{4}) - 1 = 2.833 
\]

This means, at roll 1, me accept  3, 4, 5 or 6 and roll again otherwise, which gives:

\[
E_1 = (\frac{4}{6} \times \frac{3 + 4 + 5 + 6}{4}) + (\frac{2}{6} \times 2.833) = 3.944
\]

$\therefore$ the expected reward under optimal strategy is 3.944.
\end{solution}

\begin{question}
A 10 sided die can be rolled continuously. On each roll, if the face shows a number less than 10,
that value is added to the current reward. If the face shows a 10, the game ends with 0 reward.
What is the optimal playing strategy?
\end{question}
\begin{solution}
The optimal stopping decision is conditioned on the current balance of rewards: what is the point
after which, the new expected reward stops being more than the current reward?

Let the current balance be $S$ . Let the new balance after a roll be $S + k$ for some $k$.
\[S + E[k] = S + \frac{9}{10} \times (S + 5)\]
\[E[k] \geq 0 \implies \frac{S}{10} - 4.5 \leq 0 \implies S \leq 45 \]
So, we must stop as soon as the current balance is atleast 45.

\end{solution}

\newpage
\section*{Linearity Of Expectation}
\begin{question}
A shuffled deck of 50 black and 50 red cards is laid out linearly. What is the expected number of
runs?
\end{question}
\begin{solution}
Define the indicator variables $I_j$ for $j \in \{1, 2, \dots, 100\}$ such that $I_j = 1$ if a new 
run begins at position $j$, and $I_j = 0$ otherwise.

The total number of runs $R$ is given by:
\[ R = I_1 + \sum_{j=2}^{100} I_j \]

By linearity of expectation:
\[ E[R] = E[I_1] + \sum_{j=2}^{100} E[I_j] \]

$E[I_1] = 1$ and by symmetry, $E[I_i] = 2 \times \frac{50}{100} \times \frac{50}{99}$. Hence,

\[
E[R] = 1 + 99 \times 2 \times \frac{50}{100} \times \frac{50}{99} = 1 + 50 = 51
\]
\end{solution}

\newpage
\section*{State Propogation}
\begin{question}
A fair 6 sided die is rolled until the difference of two consecutive rolls is atmost 1. What is
the expected number of rolls?
\end{question}
\begin{solution}
Let $E_i$ denote the extra expected number of rolls after this point, if the current roll was number
$i$, where $i=1,2,...6$. Let E be the required number.
\begin{equation}
E = 1 + \frac{1}{6} (E_1 + E_2\:.\:.\:.\:+ E_6)
\end{equation}
Based on how the states propogate, we have the following equations:
\[E_1 = 1 + \frac{1}{6} (E_3 + E_4 + E_5 + E_6)\]
\[E_2 = 1 + \frac{1}{6} (E_4 + E_5 + E_6)\]
\[E_3 = 1 + \frac{1}{6} (E_1 + E_5 + E_6)\]
\[E_4 = 1 + \frac{1}{6} (E_1 + E_2 + E_6)\]
\[E_5 = 1 + \frac{1}{6} (E_1 + E_2 + E_3)\]
\[E_6 = 1 + \frac{1}{6} (E_1 + E_2 + E_3 + E_4)\]
Solving these equations and substituting in (4), we get E = 3.27
\end{solution}
\begin{question}
A fair 6 sided die is rolled until the difference of two consecutive rolls is 2. What is the 
expected number of rolls?
\end{question}
\begin{solution}
Let $Ei$ denote the extra expected number of rolls needed, to achieve the result from a point 
where the current role is $i$. Let E denote the expected number of rolls to achieve the result.
\begin{equation}
E = 1 + \frac{1}{6}(E1 + E2 \:.\:.\:.\:+E6)
\end{equation}
Based on how the states propogate, we have the following equations:
\[E1 = 1 + \frac{1}{6}(E1 + E2 + 0 + E4 + E5 + E6)\]
\[E2 = 1 + \frac{1}{6}(E1 + E2 + E3 + 0 + E5 + E6)\]
\[E3 = 1 + \frac{1}{6}(0 + E2 + E3 + E4 + 0 + E6)\]
\[E4 = 1 + \frac{1}{6}(E1 + 0 + E3 + E4 + E5 + 0)\]
\[E5 = 1 + \frac{1}{6}(E1 + E2 + 0 + E4 + E5 + E6)\]
\[E6 = 1 + \frac{1}{6}(E1 + E2 + E3 + 0 + E5 + E6)\]
Solving these and substituting back in (6), we get $E = 5.67$.
\end{solution}

\begin{question}
A fair coin is tossed repeatedly until an "HT" is observed. What is the expected number of tosses?
\end{question}
\begin{solution}
Let $E_H$ = expected extra number of rolls to get a "HT" having already a H.

$E_T$ = expected extra number of rolls to get a "HT" having already a T.

$E_{HT}$ = expected number of rolls to get an HT.
\[E_{HT} = \frac{1}{2} \times (1 + E_{H}) + \frac{1}{2} \times (1 + E_{T})\]
\[E_H = \frac{1}{2} \times (1) + \frac{1}{2} \times (1 + E_H)\]
\[E_T = E_{HT}\]
Solving these, we get $E_H$ = 2 and $E_{HT} = 4$
\end{solution}
\begin{question}
A fair coin is tossed repeatedly until an "HH" is observed. What is the expected number of tosses?
\end{question}
\begin{solution}
Let $E_{HH}$ = expected number of tosses to observe HH.

$E_H$ = expected number extra tosses to get HH if the current toss is an H.

$E_T$ = expected number extra tosses to get HH if the current toss is a T.
\[E_{HH} = \frac{1}{2} \times (1 + E_H) + \frac{1}{2} \times (1 + E_T)\]
\[E_H = \frac{1}{2} \times (1) + \frac{1}{2} \times (1 + E_T)\]
\[E_T = E_{HH}\]
Solving these, we get $E_{HH} = 6$
\end{solution}
\end{document}